Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-04-28 paper/ms.tex
update ms

\author{et al}
%
%% Early experiments for Kepler rotation samples
%\author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
%\affiliation{\caltech}
%
%% ?
%%\author{Lynne~A.~Hillenbrand}
%%\affiliation{\caltech}
%
%% HIRES observations and reductions
%\author[0000-0002-0531-1073]{H. Isaacson}
%\affiliation{\berkeley}
%% HIRES observations, reductions, JUMP infrastructure
%\author[0000-0001-8638-0320]{A. W. Howard}
%\affiliation{\caltech}
%%
%\author{HIRES observers to be named}
%\affiliation{Various}

% <250 words
\begin{abstract}
  Recent studies of stars in open clusters have shown how the rotation
  rates and lithium content of FGK stars evolve during the first few
  billion years of their lives.
  %
  Here we apply this knowledge to stars observed by the primary Kepler
  mission to calculate rotation and lithium-based ages.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid;
  these rotation-based ages are $\approx$15-20$\times$ more precise
  than isochrone-based ages for our sample of FGK stars.
  %
  The independent lithium ages are consistent for XX\% of cases;
  however this method is less precise than rotation-based age dating
  for most of our sample.
  %
  Our resulting catalog of planetary ages includes
  \nplyounggyrotwosigma\ planets younger than 1\,Gyr at 2$\sigma$
  (\nplyounggyrothreesigma\ planets at 3$\sigma$).
  %
  In the broader sample of stars in the Kepler field, we find that the
  rotation-based age distribution decreases linearly from 3\,Gyr ago
  to present day, implying a linearly decreasing star formation rate.
  %
  While this parent stellar age distribution has implications for the
  star formation history of our Galaxy, it also bounds the relative
  balance of young and old planets detectable using both Kepler and
  also future wide-field transit surveys.
  %
\end{abstract}

@@

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.
Most of these exoplanets have ages comparable to that of the solar
system.
This fact leaves many gaps in our knowledge of the exact dynamical
and physical processes that produce these objects.
Finding and characterizing young ($\lesssim$1\,Gyr) exoplanets offers
one possible workaround.
The reason is that processes such as thermal cooling, atmospheric
loss, and dynamical instability occur over $\approx$10\,Myr to
$\approx$1\,Gyr timescales, and so they can leave observable imprints in
the properties of young planets (CITE theory reviews).
Individual exemplars include V1298~Tau \citep{David_2019}, an
adolescent resonant chain of four transiting puffy planets,
and HIP~67522 \citep{Rizzuto_2020}, a Jupiter-sized planet with a
Neptunian mass (CITE).
% And perhaps also TOI837, now with a uber massive core!
Population-level examples include tentative differences in the size
distribution of young vs.~old planets
(CITEVach,CITEChristianse,CiteDavid).

@@

To begin, let us imagine that the age distribution of stars within a
few hundred parsecs is uniformly distributed between 0 and 10\,Gyr
\citep[e.g.][]{Nordstrom_2004}.  
This approximiation is useful because it clarifies the rarity of young
stars:
$\approx$1\% of nearby stars have ages below 100\,Myr,
and $\approx$10\% are younger than 1\,Gyr.
Studies of exoplanets forming in disks (CITE), and of exoplanets
evolving shortly after disk dispersal (CITE CITE), are therefore
capped in their maximum achievable sample sizes by the tyranny of the
Galactic star formation rate.

@@

Despite the observational restrictions, the field of young close-in
planet discovery has seen remarkable success over the past ten years.
Pioneering discoveries with the prime Kepler mission (hereafter,
``Kepler'') focused on stars with known ages---stars in open
clusters---and searched them for transiting planets
\citep{Meibom_2013}.  Planets discovered in this way are assumed to
have the same age as the host cluster.  The resulting ages are precise
at the $\approx$10\% level.  K2 and TESS have yielded larger numbers
of young planets in clusters than Kepler, because each of these
missions has observed a larger fraction of the sky, and therefore more
stars with known ages
\citep[e.g.][]{Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Zhou_2022,Zakhozhay_2022,Wood_2023}.

A limitation of this ``cluster-first'' approach is that only
$\approx$1\% of stars within the nearest few hundred parsecs can
currently be associated with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
While there may be paths toward resolving this problem, this currently
means that there are many ``young'' (0.05-1\,Gyr) stars in the field
that are omitted by cluster-only searches.  Requiring ages precise to
a factor of three ($t/\sigma_t > 3$), the NASA Exolpanet Archive
currently reports $\approx$40 transiting planets younger than 1\,Gyr,
essentially all of which are in clusters.  However, Kepler discovered
$\approx$4{,}000 planets.  Unless there is a major discovery bias, we
would expect a larger fraction of the Kepler planets---perhaps 10\%,
or 400 planets---to be younger than 1\,Gyr, based on our order of
magnitude reasonsing.

Have these young Kepler planets already been found?  If they have, it
would probably have been by \citet{Berger_2020b_rpage},
\citet{David_2021} or \citet{Petigura_2022}.  The
\citeauthor{Berger_2020b_rpage} and \citeauthor{Petigura_2022} studies
focused on isochronal age-dating of individual stars, which is most
precise for stars whose luminosities and temperature separate them
from the main-sequence.  The \citeauthor{David_2021} study leveraged
both isochronal ages from an earlier analogous study
\citep{Fulton_Petigura_2018_cks_vii}, and, in their core analysis, a
relative gyrochrone age-ranking approach that sorted stars based on
whether they rotated faster or slower than a gyrochrone age of
$\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.  (TODO: note that
there's a lot more literature on "young" field Kepler planets than
these few sentences betray).

@@

	\caption{
    {\bf Sizes, orbital periods, and rotation-based ages of transiting
    exoplanets younger than one billion years.} Stars denote
    \nplyounggyrotwosigmanograzingnoruwe\ planets that are newly
    characterized in this work to have $t_{\rm gyro}$<1\,Gyr at
    $2$$\sigma$; circles are planets from the literature meeting the
    same age requirement, but with a heterogeneous set of age
    provenances.  Gray points are transiting planets from the NASA
    Exoplanet Archive older than 1\,Gyr.  The selection function
    matters: selecting for precise ages ($t/\sigma_{\rm t}>3$) would
    hide $\approx$ten $<$0.5\,Gyr planets, and add $\approx$twenty
    0.5-1\,Gyr planets.  For this plot, we require our Kepler stellar
    hosts to have no quality flags raised, and for the planets to
    similarly no have quality flags raised.  Weakening these
    constraints would expand the sample at the expense of reliability.
		\label{fig:rp_period_age_results}
	}
\end{figure*}

@@

L.G.B.~conceived the project, collected HIRES spectra, executed the
rotation- and lithium-based age analyses, and drafted the manuscript.
E.K.P.~contributed to the rotation-based age analysis.
L.A.H.~contributed to project design.
H.I. and A.W.H~contributed to acquisition, reduction, and analysis of
the HIRES data.
All authors assisted in manuscript revision.

>>>>

\author{and coauthors!}
%
%% Early experiments for Kepler rotation samples
%\author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
%\affiliation{\caltech}
%
%% ?
%%\author{Lynne~A.~Hillenbrand}
%%\affiliation{\caltech}
%
%% HIRES observations and reductions
%\author[0000-0002-0531-1073]{H. Isaacson}
%\affiliation{\berkeley}
%% HIRES observations, reductions, JUMP infrastructure
%\author[0000-0001-8638-0320]{A. W. Howard}
%\affiliation{\caltech}
%%
%\author{HIRES observers to be named}
%\affiliation{Various}

% <250 words
\begin{abstract}
  Recent studies of stars in open clusters have shown how the rotation
  rates and lithium content of FGK stars evolve during the first few
  billion years of their lives.
  %
  Here we apply this knowledge to stars observed by the primary Kepler
  mission to curate catalogs of rotation and lithium-based ages.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid;
  these rotation-based ages are $\approx$15-20$\times$ more precise
  than isochrone ages for our sample of FGK stars.
  %
  While the independent lithium ages are consistent with XX\% of the rotation ages,
  this method is significantly less precise than rotation-based age dating
  for most of our sample.
  %
  Our resulting catalog of planetary ages includes
  \nplyounggyrotwosigma\ planets younger than 1\,Gyr at 2$\sigma$
  (\nplyounggyrothreesigma\ planets at 3$\sigma$).
  %
  In the broader sample of stars in the Kepler field, we find that the
  rotation-based age distribution decreases linearly from 3\,Gyr ago
  to present day, implying a linearly decreasing star formation rate.
  %
  While this parent stellar age distribution has implications for the
  star formation history of our Galaxy, it also bounds the relative
  balance of young and old planets detectable using both Kepler and
  also future wide-field transit surveys.
  %
\end{abstract}

@@

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.
Most of these exoplanets have ages comparable to that of the solar
system.
This fact leaves many gaps in our knowledge of the exact dynamical and
physical processes that produce these objects.
Finding and characterizing young ($<$1\,Gyr) exoplanets offers
one possible workaround.
The reason is that processes such as thermal cooling
\citep{2007ApJ...659.1661F}, atmospheric loss
\citep{2019AREPS..47...67O}, giant impacts
\citep{2014prpl.conf..595R}, and dynamical instability
\citep{2017MNRAS.470.1750I} are expected to over $\approx$10\,Myr to
$\approx$1\,Gyr timescales, and so they may leave observable imprints
in the properties of young planets.
Individual exemplars include V1298~Tau \citep{David_2019}, an
adolescent resonant chain of four transiting puffy planets, and
HIP~67522 \citep{Rizzuto_2020}, a Jupiter-sized planet with a
Neptunian mass (Thao et al.~submitted).
% And perhaps also TOI837, now with a uber massive core!
Population-level examples include tentative differences in the size
distribution of young planets relative to their older counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

@@

To begin, let us imagine that the age distribution of stars within a
few hundred parsecs of the Sun is uniform between 0 and 10\,Gyr
\citep[e.g.][]{Nordstrom_2004}.  This approximation is useful because
it clarifies the rarity of young stars: under these assumptions,
$\approx$1\% of nearby stars have $t$$<$100\,Myr, and $\approx$10\%
have $t$$<$1\,Gyr.  Studies of forming protoplanetary
\citep{2018A&A...617A..44K}, and of exoplanets evolving shortly after
disk dispersal \citep[e.g.][]{2022MNRAS.512.5067K}, are therefore
capped in their maximum achievable sample sizes by the tyranny of the
Galactic star formation rate.

@@

Despite the observational restrictions, young close-in planet
discovery has seen remarkable success over the past decade.
This has been enabled by a combination of Kepler, K2, TESS, and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The original strategy from around 2010 to 2020 was to focus on stars
with known ages---obvious members of open clusters---and to search
them for transiting planets.  
The resulting stellar ages are precise at the $\approx$10\% level, and
planets discovered in this way are generally assumed to have the same
age as the host cluster.  
A more recent development, facilitated by the second Gaia data
release, is the idea that finding transiting planets can often be
easier than finding and characterizing the host stellar association.

Despite recent advances, the main problem with any ``cluster-first'' approach is that
only $\approx$1\% of stars within the nearest few hundred parsecs can
currently be associated with their birth cluster, even using the Gaia
data
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
In other words, the majority of nearby sub-gigayear stars are in the
field, and are omitted by cluster-only planet searches.  
%TODO FIXME: is this rhetorical point fair?
One can appreciate the problem by querying NASA's Exoplanet Archive
for transiting planets younger than 1\,Gyr.
Imposing reasonable precision requirements, e.g. either $t$$<$1\,Gyr
at 2$\sigma$, or else requiring ages precise to say S/N$=$3, gives $\approx$50 ``young'' transiting planets at the time of writing.
Most of these planets are in clusters, and have been discovered by Kepler, K2, or TESS.  
However, combined, these surveys have discovered
$\approx$5{,}000 planets.  Barring a major discovery bias, we
would expect an order of magnitude more transiting 
planets to be younger than 1\,Gyr, based on our earlier order of
magnitude reasoning.

This study aims to resolve two main issues.
First, there is the question of how wrong it is to assume a uniform age
distribution for stars in the galactic thin disk, particularly young stars.
Second, there is the question of which planets from the main Kepler
mission are actually young.

Given that Kepler's primary mission ended in 2013, one might be forgiven for
assuming that the ages of the Kepler planets are already known.
Of course, they have been studied:
\citeauthor{Berger_2020b_rpage} and \citeauthor{Petigura_2022} for instance calculated
isochronal ages for individual stars, which is most
precise for stars whose luminosities and temperatures separate them
from the main-sequence.
\citeauthor{David_2021} leveraged
both isochronal ages from an earlier analogous study
\citep{Fulton_Petigura_2018_cks_vii}, and, in their core analysis, this work used a
relative gyrochrone age-ranking approach that sorted stars based on
whether they rotated faster or slower than a gyrochrone age of
$\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.
These studies were conducted in the broader context of investigating
trends in exoplanet demographics, which have been shown to depend on
both stellar metallicity and mass \citet{Petigura_2018}, at levels
likely stronger than age (at least over gigayear timescales).
However, these studies offered limited sensitivity to the ages of the youngest
planets, due to the limitations of isochronal dating.

%      and \citet{Petigura_2022} have reported trends between small-planet
%      occurrence and both stellar metallicity and mass; these parameters
%      seem, at first glance, to be more important predictors of bulk changes
%      in the planetary population than age.  Given these known correlations,
%      it can be useful to consider population models in which variables such
%      as metallicity, mass, and age are added in succession
%      \citep[e.g.][]{Thorngren_2021}.  An alternative, and perhaps more
%      rigorous approach, is to jointly infer the number of planets per star
%      as a function of both planetary and stellar parameters, the latter
%      including age \citep[e.g.][]{2023AJ....166..209M}.

@@

	\caption{
    {\bf Sizes, orbital periods, and rotation-based ages of transiting
    exoplanets younger than one billion years.} Stars denote
    \nplyounggyrotwosigmanograzingnoruwe\ planets that are newly
    characterized in this work to have $t_{\rm gyro}$<1\,Gyr at
    $2$$\sigma$; circles are planets from the literature meeting the
    same age requirement, but with a heterogeneous set of age
    provenances.  Gray points are transiting planets from the NASA
    Exoplanet Archive older than 1\,Gyr.  If
 	one were to select for precise ages ($t/\sigma_{\rm t}>3$), this would
    hide $\approx$ten $<$0.5\,Gyr planets, and add $\approx$fifteen
    0.5-1\,Gyr planets.  For this plot, we require our Kepler stellar
    hosts to have no quality flags raised, and for the planets to
    similarly no have quality flags raised. 
		\label{fig:rp_period_age_results}
	}
\end{figure*}

@@

L.G.B.~conceived the project, collected data, 
inferred ages, and drafted the manuscript.
E.K.P.~contributed to early iterations of the rotation-based age analysis.
L.A.H.~contributed to project design.
H.I. and A.W.H~contributed to acquisition, reduction, and analysis of
the HIRES data.
All authors assisted in manuscript revision.
